---
title: "Non-parametric Tests"
author: "Adam M. Wilson"
date: "November 12, 2014"
output: ioslides_presentation
runtime: shiny
---
```{r, echo=FALSE, warning=FALSE}
library(knitr)
library(ggplot2)
```
## Shiny Presentation

This R Markdown presentation is made interactive using Shiny. The viewers of the presentation can change the assumptions underlying what's presented and see the results immediately. 

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Distributional assumptions
```{r}
mu=5
sigma=2.5

x=seq(-20,20,len=1000)
y=dnorm(x,mu,sigma)
d=cbind.data.frame(x,y)

ggplot(d,aes(x=x,y=y))+geom_line()+
  geom_segment(aes(x=mu,xend=mu,y=0,yend=dnorm(0.5,mean=mu,sd=sigma)),col="red")

n=1000
d=data.frame(x=c(rnorm(n,5,3),rnorm(n,20,5)))

curve(dnorm(x,0.37,1),-2,2)

ggplot(d,aes(x=x))+geom_histogram()#+stat_bin(100)


```



## 2-group Mann-Whitney U Test 
```{r}
gs=function(n){
  A=sample(c("A","B"),n,replace=T)
  y=rnorm(n)
  cbind.data.frame(A=A,y=y)}

#wilcox.test(y~A) 
```
where y is numeric and A is A binary factor

## Slide with Interactive Plot

```{r, echo=FALSE}
inputPanel(
  selectInput("n", label = "Number of samples:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("beta", label = "Slope:",
              min = -2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  data=gs(as.numeric(input$n))
  hist(data$y)
  
#  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
#  lines(dens, col = "blue")
})
```

## Immer data
```{r,warning=FALSE}
library(MASS)         # load the MASS package 
kable(head(immer)) 
```


## Plot
```{r}
plot(immer$Y1, immer$Y2) 
abline(0,1)
```

## test
```{r}
wilcox.test(immer$Y1, immer$Y2, paired=TRUE) 
```

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Code and Output

```{r}
summary(cars)
```


