---
title: "Introduction to R"
author: "Adam M. Wilson"
date: "February 11, 2015"
output: 
  html_document:
    toc: true
    keep_md: true
---


```{r, echo=FALSE, message=FALSE, results='hide', purl=FALSE}
## This chunk automatically generates a text .R version of this script when running within knitr.  You do not need to run this...
input  = knitr::current_input()  # filename of input document
output = paste(tools::file_path_sans_ext(input), 'R', sep = '.')
knitr::purl(input,output,documentation=2,quiet=T)
repo="adammwilson/SpatialAnalysisTutorials/blob/master/R_Introduction"
knitr::opts_chunk$set(cache=TRUE)
```

#  Introduction
This script is available:

  * [SpatialAnalysisTutorials repository](`r paste0("http://github.com/",repo)`)
  * Plain text (.R) with commented text 
  [here](`r paste0("https://raw.githubusercontent.com/adammwilson/SpatialAnalysisTutorials/master/R_Introduction/",output)`)
 


## Variables

```{r echo=FALSE}
x=1
x
````

We can also assign a vector to a variable:
```{r}
x=c(5,8,14,91,3,36,14,30)
x
```

And do simple arithmetic:
```{r}
x+2
```

Note that `R` is case sensitive, if you as for `X` instead of `x`, you will get an error
```{r,eval=FALSE}
X
```

`Error: object 'X' not found`

### Subsetting
Subset the vector using `x[ ]` notation
```{r}
x[5]
x[1:5]
```

### Using Functions

To calculate the mean, you could do it _manually_ like this

```{r}
(5+8+14+91+3+36+14+30)/8
```

Or use a function:
```{r}
mean(x)
```

Type `?functionname` to learn more about a function, e.g. `?mean`.  In RStudio, you can also search in the help panel.  There are other arguments too: `mean(x, trim = 0, na.rm = FALSE, ...)`

If you press `TAB` after a function name (such as `mean(`), it will show function arguments.

> Try to calculate the mean of `c(3,6,12,89)`.    

### Logical values

R also has standard conditional tests to generate `TRUE` or `FALSE` values.  These are often useful for filtering data (e.g. identify all values greater than 5).  The logical operators are `<`, `<=`, `>`, `>=`, `==` for exact equality and `!=` for inequality.

```{r}
  x = 50
  x > 75
 
  x == 40
 
  x >   15
 
```

And of course you can save the results as variables:
```{r}
result =  x >  3
result
```



### Generating Data

There are many ways to generate data in R such as sequences:
```{r}
seq(from=0, to=1, by=0.25)
```
and random numbers that follow a statistical distribution (such as the normal):

```{r}
a=rnorm(100,mean=0,sd=10)
## illustrate with a histogram
hist(a)
```


## Matrices
You can also use matrices (2-dimensional arrays of numbers):
```{r}
y=matrix(1:30,ncol=5)
y
```

Which behave much like vectors:
```{r}
y+2
```

and have 2-dimensional indexing:
```{r}
y[2,3]
```

## Data Frames
Data frames are similar to matrices, but more flexible.  Matrices must be all the same type (e.g. all numbers), while a data frame can include multiple data types (e.g. text, factors, numbers). Dataframes are commonly used when doing statistical modeling in R.  

```{r}
data = data.frame( x = c(11,12,14),
                   y = c(19,20,21),
                   z = c(10,9,7))
data

mean(data$z)

mean(data[["z"]])

mean(data[,3])
```

## Loading Packages

One of the best things about 
To load a package, you can simply type `library(package)` where `package` is the name of the package you want to load.  However, this only works for packages that you already have installed on your system.  To install new packages, you can use `install.packages()` or use the package manager. 

```{r message=F,warning=FALSE}
library(raster)
```

> R may ask you to choose a CRAN mirror. CRAN is the distributed network of servers that provides access to R's software.  It doesn't really matter which you chose, but closer ones are likely to be faster.  From RStudio, you can select the mirror under Tools→Options or just wait until it asks you.


If you don't have the packages above, install them in the package manager or by running `install.packages("raster")` (or use the package manager). 

## Working with Raster Data
```{r}

```



# Coda



```{r, echo=FALSE, message=FALSE, results='hide', purl=FALSE}
#input=tools::file_path_sans_ext(input)
```



