The data and the model
======================

We will only need one package for this exercise:
```{r, message=FALSE, warning=FALSE}
  library(raster)
```


Simple Normal world
-------------------

The world is a raster of 50 x 50 cells. Each cell contains value that is an independent outcome of a **Normally distributed random process with given mean and variance**:
```{r, fig.width=14, fig.height=5}
  # the parameters of the Normal distribution
  mean <- 0
  sd <- 1
  
  # random draws from the Normal distribution
  normal.world <- raster(matrix(NA, nrow=50, ncol=50))
  normal.world[] <- rnorm(2500, mean, sd)

  par(mfrow=c(1,3))
  # plotting the probability density function
  curve(dnorm(x, mean, sd), 
        from=min(normal.world[]), 
        to=max(normal.world[]), 
        main="Probability Density Function - PDF",
        xlab="x",
        ylab="Probability Density")
  # plotting the raster world
  plot(normal.world, axes=FALSE, main="x in space")  
  # plotting the empirical histogram
  hist(normal.world, xlab="x", ylab="Frequency", 
       main="Frequency histogram", col="grey")  
  
```


p(D|H): The Likelihood
----------------------
Calculation of likelihood in R is easy! The R functions ```dnorm()```, ```dpois()```, ```dunif()```, ```dbinom()```, ```dgamma()```, ```dbeta()``` etc. are exactly for that!

Example: What is the probability that x will have the observed values of 1.5, given the Normal(mean, sd) model?
```{r, fig.width=5, fig.height=5}
  mean <- 0
  sd <- 1
  x <- 1.5
  
  # this is how you calculate the likelihood:
  likelihood <- dnorm(x, mean, sd)
  likelihood
```

We can plot it:
```{r, fig.width=5, fig.height=5}  
  curve(dnorm(x, mean, sd), from=qnorm(0.001, mean, sd), 
                            to=qnorm(0.999, mean, sd),
        main=paste("x =", x, "; likelihood =", round(likelihood,3)),
        xlab="x",
        ylab="Probability Density")
  abline(v=x, col="red", lty=2)
  abline(h=likelihood, col="red", lty=2)
```


Simple Binomial world
---------------------

The world is a raster of 50 x 50 cells. Each cell contains value that is an independent outcome of a **Binomially-distributed random process with given parameters $p$ and $n$** (number of "coin tosses"):

```{r, fig.width=14, fig.height=5}
  # the parameters of the Binomial distribution 
  # Note: if n.tosses=1 then we do Bernoulli trial
  p=0.3
  n.tosses=1

  # random draws from Binomial
  binomial.world <- raster(matrix(NA, nrow=50, ncol=50))
  binomial.world[] <- rbinom(2500, size=n.tosses, p=p)

  par(mfrow=c(1,3))
  # plotting the probability mass function
  plot(0:max(binomial.world[]), 
       dbinom(0:max(binomial.world[]), size=n.tosses, p=p),
       type="h", lwd=4,
       xlab="Number of successes",
       ylab="Probability Mass",
       main="Probability Mass Function - PMF")
  # plotting the raster
  plot(binomial.world, axes=FALSE, main="Number of successes in space")  
  # plotting empirical histogram
  hist(binomial.world, 
       xlab="Number of successes", 
       ylab="Frequency", main="Frequency histogram",
       col="grey")  
```

How would you calculate likelihood of observing a successful outcome of a trial, given p=0.1?

Simple Poisson world
--------------------

The world is a raster of 50 x 50 cells. Each cell contains value that is an independent outcome of a **Poisson-distributed random process with given parameter lambda ($\lambda$)**:

```{r, fig.width=14, fig.height=5}
  # the parameter of the Poisson distribution
  lambda=1

  # random draws from the Poisson
  poisson.world <- raster(matrix(NA, nrow=50, ncol=50))
  poisson.world[] <- rpois(2500, lambda=lambda)

  par(mfrow=c(1,3))
  # plot of the probability mass function
  plot(0:max(poisson.world[]), 
       dpois(0:max(poisson.world[]), lambda),
       type="h", lwd=4,
       xlab="Number of Cases",
       ylab="Probability Mass",
       main="Probability Mass Function - PMF")
  # plotting the raster
  plot(poisson.world, axes=FALSE, 
       main="Number of Cases in space" )  
  # plotting empirical histogram
  hist(poisson.world, xlab="Number of Cases", 
       ylab="Frequency", main="Frequency histogram",
       col="grey")  
```

How would you calculate likelihood of observing a 10 cases, given that $\lambda$=5?

Introducing spatial gradient
----------------------------

Let's assume that there is an evnironmental variable that has a form of a linear gradient in space:
```{r, fig.width=5, fig.height=5}
  environment <- raster(matrix(NA, nrow=50, ncol=50))
  environment[] <- rep(1:50, each=50)
  plot(environment, main="Predictor (Environmental Variable)")
```

Now we can make the observed **Number of Cases** (the Response) to be Poisson-distributed and dependent on the environmental variable. The formal definition of this would be
$Response_i \sim  Poisson(Gradient_i)$
where $i$ is the cell-specific index.
```{r, fig.width=5, fig.height=5}
  response <- raster(matrix(NA, nrow=50, ncol=50))
  response[] <- rpois(2500, lambda=environment[])
  plot(response, main="Response (Number of Cases)")
```

Now we can plot it all together:
```{r, fig.width=14, fig.height=5}
  par(mfrow=c(1,3))
  plot(environment, main="Predictor (Environmental Variable)")
  plot(response, main="Response (Number of Cases)")
  plot(environment[], response[],
       pch=19, col="lightblue",
       xlab="Predictor", ylab="Response (Number of Cases)",
       main="The data and the model")
  abline(a=0, b=1, lwd=2, col="red")
  # the 95% prediction intervals
  lines(1:50, qpois(p=0.975, lambda=1:50), col="red")
  lines(1:50, qpois(p=0.025, lambda=1:50), col="red")
```

And here is a proof of the fact that it is not the distribution of your data that matters for the choice of the proper statistical tool:
```{r, fig.height=4, fig.width=4}
  hist(response, main="", xlab="Response (Number of Cases)",
       col="grey")
```
The data are clearly non-poisson, yet each data point was generated by a poisson-distributed random process!

