---
title: "Introduction to Raster Package"
author: "Adam M. Wilson"
date: "February 23, 2015"
output: 
  html_document:
    toc: true
    keep_md: true
---


```{r, echo=FALSE, message=FALSE, results='hide', purl=FALSE}
## This chunk automatically generates a text .R version of this script when running within knitr.  You do not need to run this...
input  = knitr::current_input()  # filename of input document
output = paste(tools::file_path_sans_ext(input), 'R', sep = '.')
knitr::purl(input,output,documentation=2,quiet=T)
repo="adammwilson/SpatialAnalysisTutorials/blob/master/R_RasterIntroduction"
knitr::opts_chunk$set(cache=TRUE)

Sys.setenv(PATH="/Library/Frameworks/GDAL.framework/Programs/:/Library/Frameworks/Python.framework/Versions/2.7/bin:/Applications/:${PATH}")
```

----

This script is available:

  * [SpatialAnalysisTutorials repository](`r paste0("http://github.com/",repo)`)
  * Plain text (.R) with commented text 
  [here](`r paste0("https://raw.githubusercontent.com/adammwilson/SpatialAnalysisTutorials/master/R_RasterIntroduction/",output)`)
 

## Starting R on Omega

Remember to `source` the .bashrc file:
```{}
source .bashrc
```


```{r}
library(raster)
library(raster,lib.loc="/lustre/scratch/client/fas/geodata/aw524/R/")
```

## Work with climate data

First set the path to the data directory.  You'll need to uncomment the line setting the directory to `lustre/...`.

```{r}
datadir="~/Downloads/bio1-9_30s_bil"
#datadir="/lustre/scratch/client/fas/geodata/aw524/data/"

outputdir="~/scratch/data"
if(!file.exists(outputdir)) dir.create(outputdir,recursive=T)
```


```{r}
file=paste0(datadir,"/bio_9.bil") 
## now use that to load the raster dataset:
tmean=raster(file)
```


## Calling outside functions
R can do almost anything, but it isn't always the fastest at it...  For example, let's compare cropping a tif file using the `raster` package and `gdal_translate`:

```{r}
system.time(r1 <<- crop(tmean, extent(-77,-74.5,6,7)))
system.time(system(paste0("gdal_translate -projwin -77 7 -74.5 6 ",file, " ",outputdir,"/cropped.tif")))
```


But there are programs that go beyond the existing functionality of R, so `system()` can be a useful way to keep your entire workflow in R and call other programs as needed.  For example, pkfilter 
```{r}
system.time(system(paste0("pkfilter -i ",file," -o test.tif -f stdev")))
```